---
title: 'R Choropleth Demo: US Arrests for Murder by State'
author: "Brian High"
date: "Nov. 10th, 2015"
output:
  html_document:
    keep_md: yes
---

This [code](https://www3.amherst.edu/~nhorton/r2/examples/advanced.R) example 
was found in the [book](https://www.crcpress.com/Using-R-and-RStudio-for-Data-Management-Statistical-Analysis-and-Graphics/Horton-Kleinman/9781482237368):

> Horton, Nicholas J, and Ken Kleinman. *Using R 
> and Rstudio for Data Management, Statistical 
> Analysis, and Graphics*. Boca Raton, Florida: 
> CRC Press, 2015.

The pages from the book which cover this example have been provided as a 
[excerpt](https://www3.amherst.edu/~nhorton/r2/choropleth.pdf)
in PDF format, available for download from the 
[author's website](https://www3.amherst.edu/~nhorton/r2/excerpt.php).

In this document, we have modified the original script by using a 
different code block for loading packages and by converting to an Rmd 
source document.

## Procedure Description

From the text (p. 193):

> We'll use the `ggmap` package to generate the plot. It builds on the `ggplot2` 
> package, which implements ideas related to the "grammar of graphics" [188]. 
> The package uses a syntax where specific elements of the plot are added to 
> the final product using special functions connected by the + symbol. Some 
> additional work is needed to merge the dataset with the state information 
> (2.3.11) and to sort the resulting dataframe (2.3.10) so that the shape 
> data for the states is plotted in order.

[188] H. Wickham. *ggplot2: Elegant Graphics for Data Analysis*. Springer, 
New York, 2009.

## Load Packages

```{r, choro0, echo=TRUE, eval=TRUE, message=FALSE, warning=TRUE}
# Install packages (if necessary)
for (pkg in c("ggmap", "dplyr")) {
    if (! suppressWarnings(require(pkg, character.only=TRUE)) ) {
        install.packages(pkg, repos="http://cran.fhcrc.org", dependencies=TRUE)
        if (! suppressWarnings(require(pkg, character.only=TRUE)) ) {
            stop(paste0(c("Can't load package: ", pkg, "!"), collapse = ""))
        }
    }
}
```

## Load Data

```{r, choro1, echo=TRUE, eval=TRUE}
# From: https://www3.amherst.edu/~nhorton/r2/examples/advanced.R
USArrests.st = mutate(USArrests, 
                      region=tolower(rownames(USArrests)),
                      murder = cut_number(Murder, 5))
us_state_map = map_data('state')
map_data = merge(USArrests.st, us_state_map, by="region")
map_data = arrange(map_data, order)
```

## Generate Choropleth Map

From the text (p. 194):

> The `scale_fill_grey()` function changes the colors from the default 
> unordered multiple colors to an ordered and print-friendly black and white 
> (see also `scale_file_brewer`). The `ggmap` package uses the Mercator 
> projection (see `coord_map()` in the `ggplot2` package and `mapproject` in 
> the `mapproject` package). Another implementation of choropleth maps can be 
> found in the `choroplethr` package.


```{r, choro2, echo=TRUE, eval=TRUE, fig.height=4}
# From: https://www3.amherst.edu/~nhorton/r2/examples/advanced.R
p0 = ggplot(map_data, aes(x=long, y=lat, group=group)) +
    geom_polygon(aes(fill = murder)) +
    geom_path(colour='black') +
    theme(legend.position = "bottom", 
          panel.background=element_rect(fill="transparent",
                                        color=NA)) +
    scale_fill_grey(start=1, end =.1) + coord_map();
plot(p0)
```