---
title: "US FBI Crime Estimates"
author: "Brian High"
date: "12/20/2015"
output:
  html_document:
    keep_md: yes
    self_contained: no
---

Explore US state crime data from the the US DOJ, FBI UCR. Plot
property crime and violent crime for the nation and by state. Use motion charts
to allow dynamic visualization of variables changing over time.

## Data Sources

United States Department of Justice, Federal Bureau of Investigation. 
(September 2015). Crime in the United States, 2014. Retrieved 
`r as.character(format(Sys.Date(), format="%B %d, %Y"))`, from UCRDATATOOL.gov.

Link to the form which can be used to select data for download as CSV:
http://www.ucrdatatool.gov/Search/Crime/State/StatebyState.cfm

Link to the FBI UCR website with more informatoun about this source:
https://www.fbi.gov/about-us/cjis/ucr/

### Data include estimates

The United States Department of Justice says: 
 
"National or state offense totals are based on data from all reporting 
agencies and estimates for unreported areas."
  
From: UCRDATATOOL.gov

### Data are not entirely comparable

* Changes in reporting from year to year, etc., make comparisons problematic.
* See numerous notes in results page (or CSV) from UCRDATATOOL.gov.

### Caution against ranking

The United States Department of Justice says:

"Many entities—news media, tourism agencies, and other groups with an interest 
in crime in our Nation—use figures from the Uniform Crime Reporting (UCR) 
Program to compile rankings of cities and counties.  These rankings, however, 
are merely a quick choice made by the data user; they provide no insight into 
the many variables that mold the crime in a particular town, city, county, 
state, region, or other jurisdiction.  Consequently, these rankings lead to 
simplistic and/or incomplete analyses that often create misleading perceptions
adversely affecting cities and counties, along with their residents."

From: http://www.ucrdatatool.gov/ranking.cfm

## Configure document rendering

Set `knitr` options for nicer rendering this document.

```{r set_options, echo=TRUE, message=FALSE}
library(knitr)
opts_chunk$set(tidy=FALSE, cache=TRUE, error=TRUE)
```

## User-defined functions

### Load Packages (function definition)

Get and load required R packages.

```{r}
# Load one or more packages into memory, installing as needed.
load_packages <- function(pkgs, repos = "http://cran.r-project.org") {
    result <- sapply(pkgs, function(pkg) { 
        if (!suppressWarnings(require(pkg, character.only = TRUE))) {
            install.packages(pkg, quiet = TRUE, repos = repos)
            library(pkg, character.only = TRUE)}})
}
```

### Download Data (function definition)

Download CSV from DOJ UCR using RCurl's `postForm()`.

```{r}
download_data <- function(output.file, url, referer, user.agent, ystart, yend) {
    # To get the CSV we must first fill out the form. Use postForm() to 
    # supply the paramters to request all data for all states. The form
    # also requires a 'user agent' string and a 'referer' [sic].
    # This information may be retrieved using, e.g., Chrome Developer Tools.
    result.string <- postForm(url,
                  StateId=52,StateId=1,StateId=2,StateId=3,StateId=4,StateId=5,
                  StateId=6,StateId=7,StateId=8,StateId=9,StateId=10,StateId=11,
                  StateId=12,StateId=13,StateId=14,StateId=15,StateId=16,
                  StateId=17,StateId=18,StateId=19,StateId=20,StateId=21,
                  StateId=22,StateId=23,StateId=24,StateId=25,StateId=26,
                  StateId=27,StateId=28,StateId=29,StateId=30,StateId=31,
                  StateId=32,StateId=33,StateId=34,StateId=35,StateId=36,
                  StateId=37,StateId=38,StateId=39,StateId=40,StateId=41,
                  StateId=42,StateId=43,StateId=44,StateId=45,StateId=46,
                  StateId=47,StateId=48,StateId=49,StateId=50,StateId=51,
                  DataType=1,DataType=2,DataType=3,DataType=4,YearStart=ystart,
                  YearEnd=yend,NextPage='Get+Table',
                  .opts = curlOptions(
                      referer = referer,
                      verbose = TRUE,
                      header = TRUE,
                      followLocation = TRUE,
                      useragent = user.agent
                  )
    )
    
    sink(output.file)
    cat(result.string)
    sink()
}
```

### Read State Data (function definition)

Get crime data for one state from the DOJ/FBI file.

```{r}
read_state_data <- function(state, datafile, datalines) { 
    # Calculate the start line for each state, 2 rows after state name appears.
    line.start <- grep(state, datalines)
    skip <- line.start[1] + 2
    
    # End line for state is first blank line after state data. This method
    # allows for state data of variable number of years (rows).
    line.end <- grep("^$", datalines[skip:length(datalines)])
    n <- line.end[1] - 2
    
    # Get column names and clean them up
    header <- strsplit(datalines[skip], split=",")
    header <- gsub(pattern = "[ -]", replacement = "_", x = header[[1]])

    state.data <- read.table(datafile, header=FALSE, sep = ",", skip = skip, 
                             nrows = n, stringsAsFactors = FALSE)
    
    # Add a column for state name so this will be retained as data is combined.
    state.data$State <- rep(state, times = nrow(state.data))
    
    # Assign column names.
    names(state.data) <- c(header, "State", recursive = TRUE)
    
    # Convert column names to lower case.
    names(state.data) <- tolower(names(state.data))
    
    return(state.data)
}
```

### Get Data (function definition)

Get crime data from the CSV file made using DOJ UCR tool.

```{r}
get_data <- function(datafile) {
    # Tidy the CSV, as it has embedded state-name header rows we need to skip.
    states <- c(state.name, "United States-Total")
    datalines <- readLines(datafile, warn = FALSE)
    st.data <- do.call("rbind", 
                       lapply(states, function(x) 
                           read_state_data(x, datafile, datalines)))
    return(st.data)
}
```

## Load packages

Load the R packages you will need, installing any that you are missing.

```{r, message=FALSE}
# Load packages
load_packages(c("datasets", "stringr", "tidyr", "ggplot2", "googleVis", 
                "R.utils", "dplyr"))
```

## Data Management

### Prepare the data folder

Create the data folder if it does not already exist.

```{r}
# Create the data folder if needed.
datadir <- "data/doj_ucr"
dir.create(file.path(datadir), showWarnings = FALSE, recursive = TRUE)
```

### Load the crime data

If you have the tidied data file, use that, or if not, use the original data
file. Failing that, download the data from the DOJ and tidy the data.

```{r, message=FALSE}
# Load the crime data into a single data frame.
tidydatafile <- paste(datadir, "us-crime-by-state.csv", sep="/")
if (file.exists(tidydatafile) == TRUE) {
    # Read the tidy data file into a data frame.
    us.crime <- read.csv(tidydatafile)
} else {
    datafile <- paste(datadir, "CrimeStatebyState.csv", sep="/")
    
    if (file.exists(datafile) == FALSE) {
        # If you don't have the original CSV then download it from DOJ/FBI.
        url.path <- "http://www.ucrdatatool.gov/Search/Crime/State"
        csv.path <- "DownCrimeStatebyState.cfm/CrimeStatebyState.csv"
        csv.url <- paste(url.path, csv.path, sep="/")
        referer <- paste(url.path, "StatebyState.cfm", sep="/")
        agent <- 'Mozilla/5.0 Chrome/67.8.9.10 Safari/567.8910'
        ystart <- 1960
        yend <- 2012
        res <- download_data(datafile, csv.url, referer, agent, ystart, yend)
    }
    
    if (file.exists(datafile) == TRUE) {
        # Tidy the original data file downloaded from the DOJ/FBI.
        us.crime <- get_data(datafile)
        
        # Write the tidy data to a CSV file for later use.
        write.csv(us.crime, tidydatafile, row.names = FALSE)
    } else {
        stop(paste("Can't find ", datafile, "!", sep=""))
    }
}
```

## Find the timespan in years

Find minimum and maximum years for dataset timespan. Use this for plot titles.

```{r}
us.crime[us.crime$state != "United States-Total",] %>% group_by(state) %>% 
    summarise(min.year=min(year), max.year=max(year)) -> min.max.years

min.states.year <- min(min.max.years$min.year)
max.states.year <- max(min.max.years$max.year)

max.us.year <- max(us.crime[us.crime$state == "United States-Total", "year"])
min.us.year <- min(us.crime[us.crime$state == "United States-Total", "year"])
```

## Property crime versus violent crime in the US

Prepare the data.

```{r}
# Subset just the two US totals of major crime groups: violent and property.
crime.groups <- c("year", "violent_crime_rate", "property_crime_rate")
us.crime.groups <- us.crime[us.crime$state == "United States-Total", 
                            crime.groups]

# Remove "_crime_rate" from variable names and capitalize the names.
names(us.crime.groups) <- capitalize(gsub("_crime_rate", "", 
                                          names(us.crime.groups)))

# Transform data structure to "long" format for plotting.
us.crime.groups.long <- gather(us.crime.groups, Type, Rate, -Year)
```

Make a stacked bar plot.

```{r}
# Make a bar plot of violent crime versus property crime in the US.
ggplot(us.crime.groups.long, aes(x = Year, y = Rate, fill=Type)) +
    geom_bar(stat='identity') + 
    ggtitle(paste("US Crime Rate per 100,000 People, ", 
                  min.us.year, "-", max.us.year, sep=""))
```

## Property crime in the US

Prepare the data.

```{r}
# Subset the property crimes.
property <- c("year", "property_crime_rate", "burglary_rate", 
              "larceny_theft_rate", "motor_vehicle_theft_rate", 
              "state", "population")
us.property.crime.by.state <- us.crime[, property]

# Shorten some crime names for shorter plot labels and capitalize them.
property <- c("Year", "Property", "Burglary", "Larceny", "Motor_Vehicle_Theft",
             "State", "Population")
names(us.property.crime.by.state) <- property

# Subset to only see US totals not by-state values.
us.property.crime <- us.property.crime.by.state[
    us.property.crime.by.state$state == "United States-Total",]

# Transform data structure to "long" form for plotting.
us.property.crime.long <- gather(us.property.crime.by.state, Type, Rate, 
                                 Burglary, Larceny, Motor_Vehicle_Theft)
```

Make a stacked bar plot.

```{r}
# Make a bar plot of property crime in the US by crime.
ggplot(us.property.crime.long, aes(x = Year, y = Rate, fill=Type)) +
    geom_bar(stat='identity') + 
    ggtitle(paste("US Property Crime Rate per 100,000 People, ", 
                  min.us.year, "-", max.us.year, sep=""))
```

## Violent crime in the US

Prepare the data.

```{r}
# Subset the violent crimes.
violent <- c("year", "murder_and_nonnegligent_manslaughter_rate", 
             "forcible_rape_rate", "robbery_rate", "aggravated_assault_rate", 
             "violent_crime_rate", "state", "population")
us.violent.crime.by.state <- us.crime[, violent]

# Shorten some crime names for shorter plot labels and capitalize them.
violent <- c("Year", "Murder", "Rape", "Robbery", "Assault", "Violent", 
            "State", "Population")
names(us.violent.crime.by.state) <- violent

# Subset to only see US totals not by-state values.
us.violent.crime <- us.violent.crime.by.state[
    us.violent.crime.by.state$state == "United States-Total",]

# Transform data structure to "long" form for plotting.
us.violent.crime.long <- gather(us.violent.crime.by.state, Type, Rate, 
                                Murder, Rape, Robbery, Assault)

```

Make a stacked bar plot.

```{r}
# Make a bar plot of violent crime in the US by crime.
ggplot(us.violent.crime.long, aes(x = Year, y = Rate, fill=Type)) +
    geom_bar(stat='identity') + 
    ggtitle(paste("US Violent Crime Rate per 100,000 People, ", 
            min.us.year, "-", max.us.year, sep=""))
```

## US property crime by state

Make a "Motion Chart" of violent crime in the US by state and crime.

```{r}
# Configure googleVis options.
op <- options(gvis.plot.tag='chart')
```

Subset the data and create a plot title variable.

```{r}
# Subset to only see US states and not the national totals.
us.property.crime.by.state <- us.property.crime.by.state[
    us.property.crime.by.state$State != "United States-Total",]

# Make a plot title to use in next heading.
plot.title <- paste("US Property Crime Rate by State per 100,000 People, ", 
    min.states.year, "-", max.states.year, sep="")
```

### `r plot.title`

```{r, results='asis'}
plot(gvisMotionChart(data = us.property.crime.by.state, idvar = "State", 
                     timevar = "Year", colorvar = "Property",
                     xvar = "Motor_Vehicle_Theft", yvar = "Burglary", 
                     sizevar="Population",
                     options=list(title=plot.title)))
```

## US violent crime by state

Make a "Motion Chart" of violent crime in the US by state and crime.

First, subset the data and create a plot title variable.

```{r}
# Subset to only see US states and not the national totals.
us.violent.crime.by.state <- us.violent.crime.by.state[
    us.violent.crime.by.state$State != "United States-Total",]

# Make a plot title to use in next heading.
plot.title <- paste("US Violent Crime Rate by State per 100,000 People, ", 
                    min.states.year, "-", max.states.year, sep="")
```

### `r plot.title`

```{r, results='asis'}
plot(gvisMotionChart(data = us.violent.crime.by.state, idvar = "State", 
                     timevar = "Year", colorvar = "Violent",
                     xvar = "Rape", yvar = "Murder", sizevar="Population",
                     options=list(title=plot.title)))
```